---
title: "Influenza Vaccination Among Pregnant U.S. Women, 2012-2020: A Qualitative Analysis"
author: "CB"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px",
options(repos = c(CRAN = "http://cran.rstudio.com")))
```

<br>

#### Link to full report PDF[here](https://github.com/CBSC73/PM566-Final_Project/blob/main/report.pdf), with references list [here](https://github.com/CBSC73/PM566-Final_Project/blob/main/references.pdf)

<br>

## __Research Question__
####     1. Among Pregnant Women in the U.S., are Influenza vaccination rates increasing or decreasing?  
####     2. Which subpopulations (U.S. state/region, age group, race/ethnicity) are driving these trends? 
<br>

## __Why Study Influenza Vaccination in Pregnant Women?__
#### +Pregnant women are particularly susceptible to the influenza virus due to their immune tolerant status in pregnancy
#### +Influenza infection during pregnancy is associated with increased risk of neurodevelopmental disorders in children, including autism and schizophrenia
#### +Vaccination during pregnancy protects another vulnerable population: newborns, especially those born preterm or with pulmonary conditions. 
<br>

## __Methods__
#### +Data source: Centers for Disease Control (CDC) website
#### +Population: Women aged 18-49 years, pregnant anytime between August 1st - survey date
#### +Data acquision: Internet panel national survey conducted at the end of Flu Season (March-April)
#### +Sample sizes (N) between 32-47K/year 
#### +Timeframe: All years between 2012-2020
<br>
<br>


```{r, Load required libraries, include=FALSE}
library(data.table)
library(dplyr)
library(tidyverse)
library(dtplyr)
library(knitr)
library(ggplot2)
library(readr)
library(plotly)
library(DT)
```

```{r load-data, include=FALSE}
if (!file.exists("https://raw.githubusercontent.com/CBSC73/PM566-Final_Project/main/Vaccination_Coverage_dataset.txt")) {
  download.file("https://raw.githubusercontent.com/CBSC73/PM566-Final_Project/main/Vaccination_Coverage_dataset.txt", 
                "Vaccination_Coverage_dataset.txt", method="libcurl", timeout = 60)  
}

vax <- data.table::fread("Vaccination_Coverage_dataset.txt", data.table=getOption("datatable.fread.datatable", TRUE)) 
```

```{r, look at variables, include=FALSE}
#Rename columns
vaxdf<-as.data.frame(vax)
vaxdf <- rename(vaxdf, Year="Survey Year/Influenza Season", Percent_Vax="Estimate (%)",Age_Ethn="Dimension Type", Geog_Type="Geography Type")

#Remove duplicates
vaxdf<-unique(vaxdf)
dim(vaxdf)

#Remove Tdap data,Puerto Rico, City of NY/Rest of NY, D.C.
vaxdf<-vaxdf%>%filter(Vaccine=="Influenza", Geography!="District of Columbia", Geography!="NY-City of New York", Geography!="NY-Rest of state", Geography!="Puerto Rico")
dim(vaxdf)

#Make variables numeric to better evaluate them
head(vaxdf)
vaxdf$Percent_Vax<-as.numeric(as.character(vaxdf$Percent_Vax))
vaxdf$Year<-as.numeric(as.character(vaxdf$Year))


#Look for missing values
dim(vaxdf)
colSums(is.na(vaxdf)) #There are 267 missing values out of 2470 (10.8%)

#Where are the missing values? Will check by variables of interest: Age, Race/Ethnicity, location (U.S., states)
vaxdfage<-vaxdf %>% filter(Age_Ethn=="Age")
vaxdfeth<-vaxdf %>% filter(Age_Ethn=="Race/Ethnicity")
vaxdfUS<-vaxdf %>% filter(Geog_Type=="National")
vaxdfSTATES<-vaxdf %>% filter(Geog_Type=="States")

#Age
colSums(is.na(vaxdfage)) #12 missing values out of 1244 (1%)
dim(vaxdfage)

#Ethnicity
colSums(is.na(vaxdfeth)) #255 missing values out of 1226 (20.1%)
dim(vaxdfeth)

#National
colSums(is.na(vaxdfUS)) #No missing values
dim(vaxdfUS)

#States
colSums(is.na(vaxdfSTATES)) #267 missing values out of 2398 (11.1%)
dim(vaxdfSTATES)

#We can see that the vast majority of missing values are Race/Ethnicity data points in the state specific data. I only plan to examine race/ethnicity on a national level however so will not try to impute NA valu1es.

```

```{r, create 2012 and 2020 datasets, include=FALSE}

#Create datasets 
vaxtrim2012<-vaxdfSTATES %>% filter(Year=="2012", Dimension=="≥18 Years")

vaxtrim2020<-vaxdfSTATES %>% filter(Year=="2020", Dimension=="≥18 Years")

#Check each dataset for NA values
colSums(is.na(vaxtrim2012))
colSums(is.na(vaxtrim2020))
#There are no missing values for Age over 18 years data in the selected years

#Remove duplicates again just in case
vaxtrim2012<-distinct(vaxtrim2012)
vaxtrim2020<-distinct(vaxtrim2020)

view(vaxtrim2012)#Only 29 states included
view(vaxtrim2020)#Only 39 states included

```

## __Data Analyses__

<br>

### __National Overview: Visualization by State__   

<br>

```{r, chloropleth set up, include=FALSE}

library(dplyr)

#Rename the Geography column "state" so it will merge 
vaxtrim2012 <- rename(vaxtrim2012, state="Geography")
vaxtrim2020 <- rename(vaxtrim2020, state="Geography")

#Get state info into dataset, merge
st_crosswalk <- tibble(state = state.name) %>%
   bind_cols(tibble(abb = state.abb))

vaxtrim2012 <- left_join(vaxtrim2012, st_crosswalk, by = "state")
vaxtrim2020 <- left_join(vaxtrim2020, st_crosswalk, by = "state")

#Create hover text
vaxtrim2012$hover<-with(vaxtrim2012, paste(state, '<br>', "Vaccination Rate(%):", Percent_Vax))
vaxtrim2020$hover<-with(vaxtrim2020, paste(state, '<br>', "Vaccination Rate(%):", Percent_Vax))

# Set up mapping details
set_map_details<- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)
```


```{r, include=FALSE}
#Create map 2012
fig2012<-plot_geo(vaxtrim2012, locationmode = 'USA-states') %>% 
  add_trace(
    z = ~Percent_Vax, 
    text = ~hover, 
    locations = ~abb,
    color = ~Percent_Vax, 
    colors = 'Greens'
  )
fig2012 <- fig2012 %>% colorbar(title = "Vaccination Rate (%)", limits = c(25,85))
fig2012 <- fig2012 %>% layout(title = paste('Influenza Vaccination Rate During Pregnancy by State, 2012'), landcolor='#e5ecf6',
    geo = set_map_details)

```


```{r, include=FALSE}
#Create map 2020
fig2020<-plot_geo(vaxtrim2020, locationmode = 'USA-states') %>% 
  add_trace(
    z = ~Percent_Vax, 
    text = ~hover, 
    locations = ~abb,
    color = ~Percent_Vax, 
    colors = 'Greens',
    showscale = FALSE
  )
fig2020 <- fig2020 %>% colorbar(limits = c(25,85)) #Make sure limits are same for both maps
fig2020<- fig2020 %>% layout(title = paste('Influenza Vaccination Rate During Pregnancy by State, 2020'), landcolor='#e5ecf6',geo = set_map_details)
```


```{r, echo=FALSE}
### Plot together 
finalfig <- subplot(fig2012, fig2020, nrows = 2) %>% 
  layout(showlegend = FALSE,
         title = paste('Influenza Vaccination Rate During Pregnancy by State in 2012 vs 2020'),
         hovermode = TRUE
         ) %>%
  colorbar(title = "Vaccination Rate (%)", limits = c(25,85))

annotations = list( 
  list( 
    x = 0.5,
    y = 0.5,
    text = "2012",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE   
  ),  
  list( 
    x = 0.5,
    y = -0.05,
    text = "2020",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ))
finalfig <- finalfig %>%layout(annotations = annotations)
finalfig
```
<br>
<br>

####  <u>Key Points</u>
####  *Overall, vaccination rates have increased between 2012-2020.
####  *The Midwest and Northeast regions appear particularly increased.
#####  _Note: Not all states included. In 2012, 29 of 50 states; in 2020, 39 of 50._

<br>

```{r, examine by region, first create region variable, include=FALSE}
vaxdfSTATES <- rename(vaxdfSTATES, state="Geography")

vaxdtSTATES<-as.data.table(vaxdfSTATES)

vaxdtSTATES <-vaxdtSTATES[, region := fifelse(state== "Alaska"| state=="Arizona" | state=="California" | state=="Colorado"| state=="Hawaii"| state=="Idaho"| state=="Montana"| state=="Nevada"| state=="New Mexico"| state=="Oregon"| state=="Utah"| state=="Washington"| state== "Wyoming", "WEST",
                fifelse(state=="Alabama"|state=="Arkansas"|state== "Delaware" | state=="Florida"|state== "Georgia" | state=="Kentucky" |state== "Louisiana"|state== "Maryland" | state=="Mississippi" |state=="North Carolina"|state=="Oklahoma"| state=="South Carolina"| state=="Tennessee" |state=="Texas"|state== "Virginia"|state== "West Virginia", "SOUTH",
                fifelse(state=="Illinois"| state=="Indiana"| state=="Iowa"|state== "Kansas"| state=="Michigan"|state== "Minnesota"| state=="Missouri"| state=="Nebraska"| state=="North Dakota"| state=="Ohio"|state=="South Dakota"|state=="Wisconsin", "MIDWEST",                        "NORTHEAST")))]
```

### __Trends by Region__ 
<br>


```{r, compare vaccine rates by region over time, echo=FALSE}
vaxdfSTATES<-as.data.frame(vaxdtSTATES)

vaxdfSTATES_regGraph<-vaxdfSTATES%>%filter(Dimension=="≥18 Years", Year==2012| Year==2020)

p<-ggplot(vaxdfSTATES_regGraph,aes(y=Percent_Vax, x=region, fill = region) 
) +   
  geom_boxplot(color="black")+
  labs(title="Influenza Vaccination Rate Among Pregnant Women in the U.S. 2012 vs 2020",  y= "Vaccination Rate (%)")+
    theme(axis.text.x=element_blank(),legend.background = element_rect(color = "black")
                                        )+
scale_fill_brewer(palette = "Set1")+
  facet_wrap(~Year, nrow=1)

ggplotly(p)
```


####  <u>Key Points</u>
####  *All U.S. regions had higher vaccination rates in 2020 than 2012.
####  *The pattern has remained consistent, lowest rates in the South, highest rates in the Northeast.
#####  _Note: Lines in each box are median values._ 
<br>

### __Delving deeper: Individual States__ 
<br>

```{r, create interactive line graphs, echo=FALSE}

gall<-vaxdfSTATES%>% filter(Dimension=="≥18 Years") %>% arrange(Year) %>% 
plot_ly(x = ~Year, y = ~Percent_Vax, color = ~state, type = "scatter", mode = "lines",
        hoverinfo = 'text',
        text = ~paste(paste(state), 
                  paste("% Vaccinated: ", Percent_Vax, sep=""), sep = "<br>")) %>% 
       layout(title = "Influenza Vaccination Among U.S Women by State, 2012-2020", plot_bgcolor='#e5ecf6', yaxis=list(title="Vaccination Rate (%)"), legend=list(title =list(text= "<br>STATE<br>")))

gall
```
######  _Note: Hover over line for more data. Double click in Right hand panel to view states individually._
<br>

####  <u>Key Points</u>
####  *Georgia has the lowest vaccination rate of all states examined. Florida is next lowest. Their rates are far below the rest of the country.
####  *Massachusetts and Rhode Island have the highest rates.

<br>

### __Examine trends by state, year, and percent vaccination rate _change_ (from 2012 to 2020): Order and Search Data.__
<br>

#### __Table 1. Change in Vaccination Rate by State, 2012 to 2020__
######      _(Numeric values are percentages)_
```{r, table create, echo=FALSE}
vaxchange<-vaxdfSTATES %>% 
  filter(Year=="2012"|Year=="2020", Dimension=="≥18 Years") %>% 
  select(state, Year, Percent_Vax) %>% 
  pivot_wider(names_from=Year, values_from=Percent_Vax)

vaxchange$Change = (vaxchange$"2020" - vaxchange$"2012")

colnames(vaxchange)<- c("State", "Year: 2012", "Year: 2020", "Change")

datatable(vaxchange, rownames = F)
```

<br>


####   <u>Key Points</u>
####   *Hawaii, Wyoming, Maryland, and New Jersey top the list of states with the largest increases in vaccine rates with an over 50% increase between 2012 to 2020. 
####   *Minnesota has a low increase percentage, but already had a vaccine rate on the high end in the 60s. Georgia had a 7% increase from 2012 but remains the state with the lowest vaccination rate.
<br>
<br>

### __How Do Age and Ethnicity Affect Influenza Vaccination Uptake?__
<br>


```{r,  data prep, include=FALSE}

vaxdtUS<-as.data.table(vaxdfUS)

#Rename terms in the Dimension variable so they're easier to work with 
vaxdtUS<-vaxdtUS[, Dimension1 := fifelse(Dimension=="≥35 Years", "35+ Years", 
                                             fifelse(Dimension=="18-24 Years", "18-24 Years", 
                                             fifelse(Dimension=="25-34 Years", "25-34 Years", 
                                             fifelse(Dimension=="≥18 Years", "18+ Years", 
                                             fifelse(Dimension=="Black, Non-Hispanic", "Black", 
                                             fifelse(Dimension=="White, Non-Hispanic", "White", 
                                             fifelse(Dimension=="Hispanic", "Hispanic", "Other")))))))]


#Make into a dataframe for graphs
vaxdfUS<-as.data.frame(vaxdtUS)
```
```{r, echo=FALSE}
#Plotly Version, AGE


agebar<-vaxdfUS%>%
   filter(Age_Ethn=="Age",Dimension1 !="18+ Years") %>%  
   plot_ly(x= ~Year, y= ~Percent_Vax, type= 'bar', color = ~Dimension1, colors = c("Sea Green", "Blue", "Purple"),
          hoverinfo = 'textage',
          textage = ~paste(paste("Year:", Year, sep=""),
                     paste("Age:", Dimension1, sep=""),
                     paste("% Vaccinated: ", Percent_Vax, sep=""),
                     sep = "<br>"))%>% 
         layout(title = "Influenza Vaccination Among U.S Women by Age Group, 2012-2020", plot_bgcolor='#e5ecf6', yaxis=list(title="Vaccination Rate (%)", range=c(0,100)), legend=list(title =list(text= "<br>Age<br>")))

agebar
```

<br>

####  <u>Key Points</u>
####  *Older women are more likely to be vaccinated than younger women.
####  *This trend has been consistent across the years. 

<br>

```{r, echo=FALSE}
#Plotly Version, ETHNICITY
ethbar<-vaxdfUS%>%
   filter(Age_Ethn=="Race/Ethnicity") %>%  
   plot_ly(x= ~Year, y= ~Percent_Vax, type= 'bar', color = ~Dimension1, colors = c("Dark Cyan","Purple", "Dark Orange", "Blue"),
          hoverinfo = 'texteth',
          textage = ~paste(paste("Year:", Year, sep=""),
                     paste("Race/Ethnicity:", Dimension1, sep=""),
                     paste("% Vaccinated: ", Percent_Vax, sep=""),
                     sep = "<br>"))%>% 
       layout(title = "Influenza Vaccination Among U.S Women by Race/Ethnic Group, 2012-2020", plot_bgcolor='#e5ecf6', yaxis=list(title="Vaccination Rate (%)", range=c(0,100)), legend=list(title =list(text= "<br>Race/Ethnic Group<br>")))

ethbar
```

<br>

####  <u>Key Points</u>
####  *Women in the "other" race/ethnicity category have the highest vaccination rates as of 2020, overtaking White women in 2014.
####  *Black women have the lowest rates of vaccination and since 2016 rates have been trending downward.

<br>
<br>

### __Vaccination Nationally (Table 2) and by State (Table 3) for Age and Race/ethnicity Subpopulations__

<br>


#### __Table 2. Nationwide: Influenza Vaccination Rate Among Pregnant U.S. Women, 2012-2020__

<br>

```{r, create interactive table of National Data, echo=FALSE}
#Create interactive table with DT package

table2<-vaxdfUS %>% select(Year, Dimension, Percent_Vax) 
colnames(table2)<- c("Year", "Age or Race/Ethnicity", "Vaccination Rate (%)")
datatable(table2, rownames = F)
```
<br>

#####  _Note: age and rate/ethnicity data are reported separately and thus cannot be examined simultaneously. Choose which groups you would like to view using the second column, or by Search box (top right)._

<br>

#### __Table 3. By State: Influenza Vaccination Rate Among Pregnant U.S. Women, 2012-2020__
```{r, create interactive table of States Data, echo=FALSE}

table3<-vaxdfSTATES %>% filter(!is.na(Percent_Vax)) %>% select(state,Year, Dimension, Percent_Vax)
colnames(table3)<- c("State", "Year", "Age or Race/Ethnicity", "Vaccination Rate (%)")
datatable(table3, rownames = F)
```
<br>
<br>

## __Conclusions__

<br>

###      *Nationwide, vaccination against Influenza virus during pregnancy is increasing since 2012
<br>

###      *All regions of the U.S. show increasing vaccination rates, though the Southern U.S. lags behind the rest of the country with low vaccination rates (below 50%).
<br>

###      *Older women are more likely to be vaccinated than younger women, though all age groups have increased their vaccine rates since 2012.
<br>

###      *Women who identify as "Other" ethnic group are the most likely to be vaccinated, and those who identify as Black are least likely to be vaccinated.
<br>
<br>

##### <img src="https://www.logo.wine/a/logo/GitHub/GitHub-Icon-White-Dark-Background-Logo.wine.svg" width="40px"> <b>View project Github Repository [here](https://github.com/CBSC73/PM566-Final_Project)</b>
